---
title: "Bayes Rules Ch 5 Homework"
subtitle: "Exercises: 5.5, 5.6, 5.9, 5.10"
author: "Matt McGehee"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  html:
    embed-resources: true
editor: source
---

```{r}
packages <- c("tidyr","dplyr","bayesrules", "ggplot2")

needs_installing <- packages[!packages %in% installed.packages()[,"Package"]]
if(length(needs_installing)) {
  install.packages(needs_installing)
}

for(pkg in packages){
  if(!pkg %in% (.packages())){
    library(pkg, character.only = TRUE)
  }
}
```

## Exercise 5.5 (Text messages)
Let random variable λ represent the rate of text messages people receive in an hour. At first, you believe that the typical number of messages per hour is 5 with a standard deviation of 0.25 messages.
  a.  Tune and plot an appropriate $Gamma(s,r)$ prior model for λ.
  b.  What is the prior probability that the rate of text messages per hour is larger than 10? Hint: learn about $pgamma()$.

<br>
Given E[λ]=5 messages per hour and var[λ] = $0.25^2$, we can find r by dividing 
the mean by the variance: $r = \frac{5}{0.25^2} = 80$.
<br>
The shape $s$ is the mean multiplied by the rate: $s=5\times80=400$
<br>
```{r}
plot_gamma(400, 80) +
  ggtitle("Gamma(400, 80) Prior for λ") +
  xlab("λ (messages per hour)") +
  ylab("Density")
```


Cumulative distribution when $λ > 10$, using $pgamma()$:
```{r}
(p_gamma <- pgamma(10, shape = 400, rate = 80))
```
Prior probability is 1 minus the cumulative distribution,
so the $P(\lambda > 10)$ is:
```{r}
(prior_prob <- 1 - p_gamma)
```

---

## Exercise 5.6 (Text messages with data) 
Continuing with Exercise 5.5, you collect data from six friends. They received 7, 3, 8, 9, 10, 12 text messages in the previous hour.
  a.  Plot the resulting likelihood function of λ
  b.  Plot the prior pdf, likelihood function, and the posterior pdf of λ
  c.  Use $summarize_gamma_poisson()$ to calculate descriptive statistics for the prior and the posterior models of λ
  d.  Comment on how your understanding about λ changed from the prior (in the previous exercise) to the posterior based on the data you collected from your friends
  
<br>
### a. Plot Poisson Likelihood
```{r}
plot_poisson_likelihood(y = c(7,3,8,9,10,12), lambda_upper_bound = 15)
```

### b. Plot Prior PDF, Likelihood Func., & Posterior PDF

```{r}
plot_gamma_poisson(shape = 400, rate = 80, sum_y = 49, n = 6)
```

### c. Descriptive Stats
```{r}
(summarize_gamma_poisson(shape = 400, rate = 80, sum_y = 49, n = 6))
```

### d. Comments
<br>
The likelihood function has pulled the Posterior PDF for λ to the right, indicating that the E[λ] is slightly to the right of our prior belief. The peak of the posterior is slightly taller as well, which is good, as this indicates a smaller variance in our posterior distribution from the added observations.
<br>

---


