---
title: "Group 2 Assignment 3"
subtitle: "Hotel & Resort Cancellations"
author:
  - Jennifer Bercera
  - Kristina Diaz-Kerkado
  - Megan Marsh
  - Matt McGehee
  - Jegg Ononiewu
date-format: long
format:
  revealjs:
    theme: dark
    css: styles.css
course: Applied Bayesian Analysis
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false

# Load necessary libraries

packages <- c("tidyr","dplyr", "ggpubr", "ggplot2", "gt", "patchwork", "bayesrules")

needs_installing <- packages[!packages %in% installed.packages()[,"Package"]]
if(length(needs_installing)) {
  install.packages(needs_installing, repos = "https://cran.rstudio.com/")
}

for(pkg in packages){
  if(!pkg %in% (.packages())){
    library(pkg, character.only = TRUE)
  }
}
```

```{r}
#| echo: false
# Load the data
data(hotel_bookings)
# head(hotel_bookings)
```

## Data Description 

```{r}
# # Convert `is_canceled` to numeric
# hotel_bookings <- hotel_bookings %>%
#   mutate(is_canceled = as.numeric(as.character(is_canceled)))
# 
# # Aggregate cancellations per year and hotel type
# cancellations_by_year <- hotel_bookings %>%
#   group_by(arrival_date_year, hotel) %>%
#   summarise(total_cancellations = sum(is_canceled, na.rm = TRUE)) %>%
#   ungroup()
# 
# # Plot cancellations across years, stratified by hotel type
# total_cancellations_plot <- ggplot(cancellations_by_year, aes(x = arrival_date_year, y = total_cancellations, color = hotel, group = hotel)) +
#   geom_line(size = 1) +
#   geom_point(size = 2) +
#   labs(title = "Hotel Cancellations Across Years",
#        x = "Year",
#        y = "Total Cancellations",
#        color = "Hotel Type") +
#   theme_minimal()

# Convert `is_canceled` to numeric
# hotel_bookings <- hotel_bookings %>%
#   mutate(is_canceled = as.numeric(as.character(is_canceled)))
# 
# # Aggregate cancellations per year and hotel type
# bookings_summary <- hotel_bookings %>%
#   group_by(arrival_date_year, hotel) %>%
#   summarise(
#     total_bookings = n(),
#     cancellations = sum(is_canceled, na.rm = TRUE),
#     kept_bookings = total_bookings - cancellations
#   ) %>%
#   ungroup()
# 
# # Reshape data to long format but exclude total_bookings
# bookings_long <- bookings_summary %>%
#   select(-total_bookings) %>%  # Remove total_bookings as it's redundant
#   pivot_longer(cols = c("kept_bookings", "cancellations"),
#                names_to = "Booking_Status", values_to = "Count")
# 
# # Rename levels for better readability
# bookings_long$Booking_Status <- factor(bookings_long$Booking_Status, 
#                                        levels = c("cancellations", "kept_bookings"),
#                                        labels = c("Canceled", "Kept"))
# 
# # Create the correct stacked bar chart
# stacked_comparison_plot <- ggplot(bookings_long, aes(x = as.factor(arrival_date_year), y = Count, fill = Booking_Status)) +
#   geom_bar(stat = "identity", position = "stack") +
#   facet_wrap(~hotel) + # Separate charts for each hotel type
#   labs(title = "Hotel Bookings: Cancellations vs. Kept Bookings",
#        x = "Year",
#        y = "Number of Bookings",
#        fill = "Booking Status") +
#   theme_minimal() +
#   scale_fill_manual(values = c("#C44D58", "#56B4E9"))  # Brownish-orange and deep teal


```

```{r}
# Ensure `is_canceled` is numeric
hotel_bookings <- hotel_bookings %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

# Aggregate cancellations per year and hotel type
bookings_summary <- hotel_bookings %>%
  group_by(arrival_date_year, hotel) %>%
  summarise(
    total_bookings = n(),
    cancellations = sum(is_canceled, na.rm = TRUE),
    kept_bookings = total_bookings - cancellations
  ) %>%
  ungroup()

# Reshape data to long format but exclude total_bookings
bookings_long <- bookings_summary %>%
  select(-total_bookings) %>%  # Remove total_bookings as it's redundant
  pivot_longer(cols = c("kept_bookings", "cancellations"),
               names_to = "Booking_Status", values_to = "Count")

# Rename levels for better readability
bookings_long$Booking_Status <- factor(bookings_long$Booking_Status, 
                                       levels = c("kept_bookings", "cancellations"),
                                       labels = c("Kept", "Canceled"))

# Aggregate cancellations per year and hotel type for line plot
cancellations_by_year <- hotel_bookings %>%
  group_by(arrival_date_year, hotel) %>%
  summarise(total_cancellations = sum(is_canceled, na.rm = TRUE)) %>%
  ungroup()

# Combined plot: Stacked bar chart with line graph overlay
combined_plot <- ggplot() +

  # Base: Stacked bar chart
  geom_bar(data = bookings_long, 
           aes(x = as.factor(arrival_date_year), y = Count, fill = Booking_Status), 
           stat = "identity", position = "stack") +
  
  # Overlay: Line graph for cancellations
  geom_line(data = cancellations_by_year, 
            aes(x = as.factor(arrival_date_year), y = total_cancellations, group = hotel, color = hotel), 
            size = 1, linetype = "dashed") +
  
  geom_point(data = cancellations_by_year, 
             aes(x = as.factor(arrival_date_year), y = total_cancellations, color = hotel), 
             size = 3) +

  # Facet by hotel type
  facet_wrap(~hotel) +

  # Labels and styling
  labs(title = "Hotel Bookings: Cancellations vs. Kept Bookings (with Cancellation Trends)",
       x = "Year",
       y = "Number of Bookings",
       fill = "Booking Status",
       color = "Hotel Type") +

  theme_minimal() +
  
  scale_fill_manual(values = c("#C44D58", "#56B4E9")) +  # Colors for stacked bars
  scale_color_manual(values = c("#ac1fdb", "#0530f2"))   # Colors for line graph

# Display the combined plot
combined_plot
```



## Hotel Bookings {.smaller}

Data Preparation
```{r}
# load the hotel_bookings dataset
data("hotel_bookings")

# filter data for city hotels
city_hotels <- hotel_bookings %>%
  filter(hotel == "City Hotel") %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

# cancellations and total bookings for city hotels
city_totals <- city_hotels %>%
  summarise(
    total_bookings = n(),
    total_cancellations = sum(is_canceled)
  )

# display data summary as a cleaner table
library(gt)
city_totals %>%
  gt() %>%
  tab_header(
    title = "Summary of City Hotel Data",
    subtitle = "Bookings and Cancellations"
  ) %>%
  cols_label(
    total_bookings = "Total Bookings",
    total_cancellations = "Total Cancellations"
  )
```

Prior and Posterior Distributions

```{r}
# define prior parameters
alpha_prior <- 8.35  # Reflects the 41.73% prior expectation
beta_prior <- 11.65
```

Update the Posterior

```{r}
# posterior calculation
y <- city_totals$total_cancellations  # Number of cancellations (successes)
n <- city_totals$total_bookings       # Total bookings (trials)

alpha_posterior <- alpha_prior + y
beta_posterior <- beta_prior + n - y

# display posterior parameters in a clear format
cat("Posterior Distribution Parameters:\n")
cat(paste0("Alpha (Successes): ", round(alpha_posterior, 2), "\n"))
cat(paste0("Beta (Failures): ", round(beta_posterior, 2), "\n"))
```

Visualize the Prior and Posterior

```{r}
# plot prior distribution
prior_plot <- plot_beta(alpha_prior, beta_prior) + 
  theme_minimal() +
  ggtitle("Prior: Beta (8.35, 11.65)") +
  theme_bw()

# plot posterior distribution
posterior_plot <- plot_beta(alpha_posterior, beta_posterior) + 
  theme_minimal() +
  ggtitle(paste0("Posterior: Beta (", alpha_posterior, ", ", beta_posterior, ")")) +
  theme_bw()

# use patchwork to combine plots side-by-side
combined_plot <- prior_plot + posterior_plot +
  plot_annotation(
    title = "Prior vs Posterior Distributions for City Hotels",
    subtitle = "The posterior distribution is updated using observed data",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
  )
combined_plot
```


## Resort Bookings {.smaller}

:::{.panel-tabset}

### Data

```{r}
data("hotel_bookings")

hotel_bookings <- hotel_bookings %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

resort_cancellations <- hotel_bookings %>%
  filter(hotel == "Resort Hotel") %>%
  group_by(arrival_date_year) %>%
  summarize(
    total_bookings = n(),
    total_cancellations = sum(is_canceled)
  ) %>%
  mutate(
    cancellation_rate = total_cancellations / total_bookings,
    arrival_date_year = as.character(arrival_date_year)
  )

totals_row <- resort_cancellations %>%
  summarize(
    arrival_date_year = "Total",  # Character string
    total_bookings = sum(total_bookings),
    total_cancellations = sum(total_cancellations),
    cancellation_rate = total_cancellations / total_bookings
  )

resort_cancellations <- bind_rows(resort_cancellations, totals_row)

resort_cancellations %>%
  gt() %>%
  tab_header(title = "Resort Hotel Cancellations") %>%
  cols_label(
    arrival_date_year = "Year",
    total_bookings = "Total Bookings",
    total_cancellations = "Total Cancellations",
    cancellation_rate = "Cancellation Rate"
  ) %>%
  fmt_number(columns = c(total_bookings, total_cancellations), decimals = 0) %>%
  fmt_percent(columns = cancellation_rate, decimals = 1) %>%
  tab_options(table.width = px(960)) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(rows = arrival_date_year == "Total")
  )

```

### Prior
After reading an article on resort cancellations and our own personal histories,
we held a prior belief that about 20% percent of bookings get cancelled.

:::{.columns}

::: {.column width="50%"}
$$
\begin{aligned}
    E[\pi] &= \frac{\alpha}{\alpha + \beta} = 0.20 \\
    \alpha &= 0.20 \times (\alpha + \beta) \\
    \beta &= (1 - 0.20) \times (\alpha + \beta) \\
          &= 0.80 \times (\alpha + \beta)
\end{aligned}
$$
:::
::: {.column width="50%"}
```{r}
#| label: beta-1-1-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta(5, 20)
```
:::
:::

### Posterior
$$
\begin{aligned}
  \pi \mid (Y = y) &\sim \text{Beta}(\alpha + y, \beta + n - y) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(5 + 87, 20 + 358 - 87) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(92, 291)
\end{aligned}
$$

```{r}
#| label: beta-binomial-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta_binomial(alpha = 5, beta = 20, y = 87, n = 358)
```

### Prior vs Posterior

```{r}
prior_plot <- plot_beta(5,20) + theme_bw() + ggtitle("Prior: Beta (5, 20)")
posterior_plot <- plot_beta(92, 291) + theme_bw() + ggtitle("Posterior: Beta (92, 291)")

ggarrange(prior_plot, posterior_plot)
```


### Thoughts

- Initially assumed **Beta(1,1)** (complete uncertainty).
- After research, believed cancellations were **low** (~20%):
  
$$
  E[\pi] = \frac{\alpha}{\alpha + \beta} = 0.20
$$

- Set weak prior: ($\alpha + \beta = 25$) (low confidence).
- Likelihood dominated posterior due to large data influence.

:::

## Combined Resort and Hotel Bookings {.smaller}

:::{.panel-tabset}

### Data

```{r}
#| echo: false

resort_cancellations <- hotel_bookings %>%
  filter(hotel == "Resort Hotel") %>%
  group_by(arrival_date_year) %>%
  summarize(
    total_bookings = n(),
    total_cancellations = sum(is_canceled)
  ) %>%
  mutate(
    cancellation_rate = total_cancellations / total_bookings,
    arrival_date_year = as.character(arrival_date_year)  
  )


totals_row <- resort_cancellations %>%
  summarize(
    arrival_date_year = "Total",  
    total_bookings = sum(total_bookings),
    total_cancellations = sum(total_cancellations),
    cancellation_rate = total_cancellations / total_bookings
  )


resort_cancellations <- bind_rows(resort_cancellations, totals_row)

resort_cancellations %>%
  gt() %>%
  tab_header(title = "Resort Hotel Cancellations") %>%
  cols_label(
    arrival_date_year = "Year",
    total_bookings = "Total Bookings",
    total_cancellations = "Total Cancellations",
    cancellation_rate = "Cancellation Rate"
  ) %>%
  fmt_number(columns = c(total_bookings, total_cancellations), decimals = 0) %>%
  fmt_percent(columns = cancellation_rate, decimals = 1) %>%
  tab_options(table.width = pct(100)) %>%  
  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_body(rows = arrival_date_year == "Total")
  )

```

```{r}
#| echo: false

city_cancellations <- hotel_bookings %>%
  filter(hotel == "City Hotel") %>%
  group_by(arrival_date_year) %>%
  summarize(
    total_bookings = n(),
    total_cancellations = sum(is_canceled)
  ) %>%
  mutate(
    cancellation_rate = total_cancellations / total_bookings,
    arrival_date_year = as.character(arrival_date_year)  
  )


totals_row <- city_cancellations %>%
  summarize(
    arrival_date_year = "Total",  
    total_bookings = sum(total_bookings),
    total_cancellations = sum(total_cancellations),
    cancellation_rate = total_cancellations / total_bookings
  )


city_cancellations <- bind_rows(city_cancellations, totals_row)


city_cancellations %>%
  gt() %>%
  tab_header(title = "City Hotel Cancellations") %>%
  cols_label(
    arrival_date_year = "Year",
    total_bookings = "Total Bookings",
    total_cancellations = "Total Cancellations",
    cancellation_rate = "Cancellation Rate"
  ) %>%
  fmt_number(columns = c(total_bookings, total_cancellations), decimals = 0) %>%
  fmt_percent(columns = cancellation_rate, decimals = 1) %>%
  tab_options(table.width = pct(100)) %>%  
  tab_style(
    style = cell_text(weight = "bold"),  # Ensure correct style formatting
    locations = cells_body(rows = arrival_date_year == "Total")
  )
```

### Beliefs

Based on research and past data trends, we establish the following prior
beliefs for cancellation rates: Resort Hotels: We assume a prior
cancellation rate of 20% based on industry trends and historical data,
as resort bookings tend to be more stable due to vacation planning and
reservations. City Hotels: We assume a higher prior cancellation.
rate of 30%, given the more dynamic nature of city travel, business trip
adjustments, and shorter booking windows, leading to increased.
cancellations.

### Prior

:::{.columns layout-ncol=2 .tiny-text}

::: {.column width="50%"}

**Resort**
$$
 \begin{aligned}
    E[\pi] &= \frac{\alpha}{\alpha + \beta} = 0.20 \\
    \alpha &= 0.20 \times (\alpha + \beta) \\
    \beta &= (1 - 0.20) \times (\alpha + \beta) \\
          &= 0.80 \times (\alpha + \beta)
\end{aligned}
$$

```{r}
#| label: beta-4-16-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: left

plot_beta(4, 16)
```
:::

::: {.column width="50%"}

**City**
$$
\begin{aligned}
    E[\pi] &= \frac{\alpha}{\alpha + \beta} = 0.30 \\
    \alpha &= 0.30 \times (\alpha + \beta) \\
    \beta &= (1 - 0.30) \times (\alpha + \beta) \\
          &= 0.70 \times (\alpha + \beta)
\end{aligned}
$$

```{r}
#| label: beta-6-4-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: left

plot_beta(6, 14)
```
:::
:::

### Likliehood

:::{.columns layout-ncol=2 .tiny-text}

::: {.column width="50%"}
**Resort**
$$
\begin{aligned}
  \pi \mid (Y = y) &\sim \text{Beta}(\alpha + y, \beta + n - y) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(4 + 87, 16 + 358 - 87) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(91, 287)
\end{aligned}
$$

```{r}
#| label: beta-binomial-4-16-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta_binomial(alpha = 4, beta = 16, y = 87, n = 358)
```
:::

::: {.column width="50%"}

**City:**
$$
\begin{aligned}
  \pi \mid (Y = y) &\sim \text{Beta}(\alpha + y, \beta + n - y) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(6 + 87, 14 + 358 - 87) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(93, 285)
\end{aligned}
$$
```{r}
#| label: beta-binomial-plot-3
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta_binomial(alpha = 6, beta = 14, y = 87, n = 358)
```
:::
:::

### Resort Comparison

```{r}
prior_plot<-plot_beta(4,16)+ theme_bw() + ggtitle("Prior: Beta (4, 16)")
posterior_plot<-plot_beta(91, 287) + theme_bw() + ggtitle("Posterior: Beta (91, 287)")

ggarrange(prior_plot, posterior_plot)
```

### City Comparison
```{r}
prior_plot<-plot_beta(6,14)+ theme_bw() + ggtitle("Prior: Beta (6, 14)")
posterior_plot<-plot_beta(93, 285) + theme_bw() + ggtitle("Posterior: Beta (93, 28)")

ggarrange(prior_plot,posterior_plot)
```

:::

## Posterior Models - Comparison {.smaller}

```{r}
#| echo: false
#| output: false

alpha_hotel <- 287.35
beta_hotel  <- 374.65

alpha_resort <- 92
beta_resort  <- 291

(ci_hotel <- qbeta(c(0.025, 0.975), alpha_hotel, beta_hotel))
(ci_resort <- qbeta(c(0.025, 0.975), alpha_resort, beta_resort))
```


| Booking Type | Posterior Summary |
|--------------|-------------------|
| **Hotel**    | $\theta_{\text{hotel}} \sim \text{Beta}(287.35,\ 374.65)$<br>$\mathbb{P}(0.40 \leq \theta_{\text{hotel}} \leq 0.47) \approx 0.95$ |
| **Resort**   | $\theta_{\text{resort}} \sim \text{Beta}(92,\ 291)$<br>$\mathbb{P}(0.20 \leq \theta_{\text{resort}} \leq 0.28) \approx 0.95$ |

  - The posterior models closely align with the observed cancellation trends in the data, likely due to:
    - the number of data observations dominating weak prior beliefs, 
    - strong prior beliefs closely aligning with the data, or 
    - a combination of both.
