---
title: "Group 2 Assignment 3"
subtitle: "Hotel & Resort Cancellations"
author:
  - Jennifer Bercera
  - Kristina Diaz-Kerkado
  - Megan Marsh
  - Matt McGehee
  - Jegg Ononiewu
date-format: long
format:
  revealjs:
    theme: dark
course: Applied Bayesian Analysis
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries

packages <- c("tidyr","dplyr","bayesrules", "ggpubr", "ggplot2", "gt")

needs_installing <- packages[!packages %in% installed.packages()[,"Package"]]
if(length(needs_installing)) {
  install.packages(needs_installing)
}

for(pkg in packages){
  if(!pkg %in% (.packages())){
    library(pkg, character.only = TRUE)
  }
}
```

```{r}
#| echo: false
# Load the data
data(hotel_bookings)
# head(hotel_bookings)
```

## Data Description 

```{r}
# Convert `is_canceled` to numeric
hotel_bookings <- hotel_bookings %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

# Aggregate cancellations per year and hotel type
cancellations_by_year <- hotel_bookings %>%
  group_by(arrival_date_year, hotel) %>%
  summarise(total_cancellations = sum(is_canceled, na.rm = TRUE)) %>%
  ungroup()

# Plot cancellations across years, stratified by hotel type
ggplot(cancellations_by_year, aes(x = arrival_date_year, y = total_cancellations, color = hotel, group = hotel)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Hotel Cancellations Across Years",
       x = "Year",
       y = "Total Cancellations",
       color = "Hotel Type") +
  theme_minimal()
```


## Slide 3

```{r}
# city_hotels <- hotel_bookings %>%
#   filter(hotel == "City Hotel") %>%
#   mutate(is_canceled = as.numeric(as.character(is_canceled)))
# 
# ggplot(city_hotels %>% filter(is_canceled == 1), aes(x = arrival_date_year)) +
#   geom_line(stat = "count") +
#   labs(x = "Year", y = "Number of Cancellations", title = "City Hotel Cancellations per Year") +
#   theme_bw()
```

## Resort Bookings {.smaller}

```{r}
# total_cancellations <- resort_cancellations$total_cancellations
# total_bookings <- resort_cancellations$total_bookings
# 
# alpha_prior <- 1
# beta_prior <- 1
# 
# plot_beta(alpha_prior, beta_prior)
# 
# alpha_post <- alpha_prior + total_cancellations
# beta_post <- beta_prior + total_bookings - total_cancellations
# 
# plot_beta_binomial(
#   alpha = alpha_prior,
#   beta = beta_prior,
#   y = total_cancellations,
#   n = total_bookings,
#   prior = TRUE,
#   likelihood = TRUE,
#   posterior = TRUE
# )

#plot_binomial_likelihood()

```

:::{.panel-tabset}

### Data

```{r}
resort_cancellations <- hotel_bookings %>%
  filter(hotel == "Resort Hotel") %>%
  group_by(arrival_date_year) %>%
  summarize(
    total_bookings = n(),
    total_cancellations = sum(is_canceled)
  ) %>%
  mutate(
    cancellation_rate = total_cancellations / total_bookings,
    arrival_date_year = as.character(arrival_date_year)
  )

totals_row <- resort_cancellations %>%
  summarize(
    arrival_date_year = "Total",  # Character string
    total_bookings = sum(total_bookings),
    total_cancellations = sum(total_cancellations),
    cancellation_rate = total_cancellations / total_bookings
  )

resort_cancellations <- bind_rows(resort_cancellations, totals_row)

resort_cancellations %>%
  gt() %>%
  tab_header(title = "Resort Hotel Cancellations") %>%
  cols_label(
    arrival_date_year = "Year",
    total_bookings = "Total Bookings",
    total_cancellations = "Total Cancellations",
    cancellation_rate = "Cancellation Rate"
  ) %>%
  fmt_number(columns = c(total_bookings, total_cancellations), decimals = 0) %>%
  fmt_percent(columns = cancellation_rate, decimals = 1) %>%
  tab_options(table.width = px(960)) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(rows = arrival_date_year == "Total")
  )

```

### Prior
After reading an article on resort cancellations and our own personal histories,
we held a prior belief that about 20% percent of bookings get cancelled.

:::{.columns}

::: {.column width="50%"}
$$
\begin{aligned}
    E[\pi] &= \frac{\alpha}{\alpha + \beta} = 0.20 \\
    \alpha &= 0.20 \times (\alpha + \beta) \\
    \beta &= (1 - 0.20) \times (\alpha + \beta) \\
          &= 0.80 \times (\alpha + \beta)
\end{aligned}
$$
:::
::: {.column width="50%"}
```{r}
#| label: beta-1-1-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta(5, 20)
```
:::
:::
### Posterior
$$
\begin{aligned}
  \pi \mid (Y = y) &\sim \text{Beta}(\alpha + y, \beta + n - y) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(5 + 87, 20 + 358 - 87) \\
  \pi \mid (Y = 87) &\sim \text{Beta}(92, 291)
\end{aligned}
$$

```{r}
#| label: beta-binomial-plot
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

plot_beta_binomial(alpha = 5, beta = 20, y = 87, n = 358)
```

<!-- ### Beta-Binomial Summary -->

<!-- ```{r} -->
<!-- summarize_beta_binomial(alpha = 5, beta = 20, y = 87, n = 358) %>% -->
<!--   gt() %>% -->
<!--   tab_header(title = "Beta-Binomial Summary") %>% -->
<!--   fmt_number(columns = c(mean, mode, var, sd), decimals = 4) %>% -->
<!--   tab_options(table.width = px(960)) -->
<!-- ``` -->

### Prior vs Posterior

```{r}
prior_plot<-plot_beta(5,20)+ theme_bw() + ggtitle("Prior: Beta (5, 20)")
posterior_plot<-plot_beta(92, 291) + theme_bw() + ggtitle("Posterior: Beta (92, 291)")

ggarrange(prior_plot,posterior_plot)
```


### Thoughts

- Initially assumed **Beta(1,1)** (complete uncertainty).
- After research, believed cancellations were **low** (~20%):
  
$$
  E[\pi] = \frac{\alpha}{\alpha + \beta} = 0.20
$$

- Set weak prior: ($\alpha + \beta = 25$) (low confidence).
- Likelihood dominated posterior due to large data influence.

:::

## Jegg

```{r}
# hotels <- hotel_bookings %>%
#   mutate(is_canceled = as.numeric(as.character(is_canceled)))
# 
# ggplot(hotels %>% filter(is_canceled == 1), aes(x = arrival_date_year)) +
#   geom_line(stat = "count") +
#   labs(x = "Year", y = "Number of Cancellations", title = "City Hotel Cancellations per Year") +
#   theme_bw()
```

## Posterior Comparison


